---
title: "Adv Data Programming with R - HW 1"
author: "Sathvik Srinath Gaurav - 24213364"
format: 
  pdf:
    geometry:                  
      - top = 0.5in
      - bottom = 0.5in
      - left = 0.5in
      - right = 0.5in
      - a4paper
    include-in-header: 
      text: |
        \usepackage{xcolor}
editor: visual
execute: 
  error: true                        
  warning: false
---

# Introduction to Dataset

-   The **GaltonFamilies** dataset comes from a 19th-century heredity study by *Sir Francis Galton*.
-   It includes height data from \textcolor{red}{204 families of 934 individuals}, with father, mother, and multiple children in each.
-   Each child is identified by gender and birth order, and their height is recorded.
-   This dataset helps us explore how parent height influences child height.

::: {.callout-note title="Galton’s Approach"}
-   Galton’s goal was to investigate how parental characteristics influence their offspring, particularly focusing on height as a measurable inherited trait.
-   Galton created a “mid-parent height” by adjusting and averaging the parent's heights to study inheritance patterns.\
    $midparentHeight = \frac{(father + 1.08 * mother)}{2.08}$
:::

```{r}
#| label: load_packages
#| echo: false

# Load required package
library(HistData)
library(ggplot2)
library(dplyr)
library(kableExtra)
```

# Summary Statistics

Using the **summary()** function on the GaltonFamilies dataset, it gives a overview of the key statistics, such as the \textcolor{blue}{minimum, maximum, mean, and quartiles} for each variable.

```{r}
#| label: load_data
#| echo: false

# Load the data
data("GaltonFamilies")

summary(GaltonFamilies)
```

-   Most fathers are between 68–71 inches tall.
-   Most children are around 66–70 inches tall.
-   Almost equal numbers of male and female children.
-   On average, each family had about 6 children, with some families having up to 15.

\
Mid-parent height is calculated and summary statistics is provided, which includes the means and standard deviations of both child and parent heights.

```{r}
#| label: summarising
#| echo: false
#| warning: false
#| message: false

# Combined parent height variable (average of father and mother)
GaltonFamilies <- GaltonFamilies %>%
  mutate(midparentHeight = (father + 1.08 * mother) / 2.08)

# Summary statistics
summary_stats <- GaltonFamilies %>%
  summarise(
    mean_child = mean(childHeight),
    mean_parent = mean(midparentHeight),
    sd_child = sd(childHeight),
    sd_parent = sd(midparentHeight)
  )

summary_stats %>%
  kable(digits = 2, caption = "Summary Statistics of Parent and Child Heights")
```

-   On average, the children in the dataset are about 66.75 inches tall.
-   The average mid-parent height (weighted average of father's and mother's height) is 66.54 inches, which is comparatively similar to average child height.
-   More variation in child height compared to parents.

# Analysis - Mid-Parent vs Child Height

-   For exploring, we analyze the relationship between mid-parent height (a weighted average of the father's and mother's heights) and the child's height.
-   The mid-parent height places an important key as it accounts for differences in average height between men and women, making it a balanced estimate of expected genetic influence. By plotting child height against this mid-parent value, we can visually inspect the strength and direction of inheritance.

```{r}
#| label: plot
#| echo: false
#| warning: false
#| message: false
#| fig-cap: " Relationship between mid-parent height (weighted average of father and mother’s height) and child height, based on Galton’s 19th-century heredity study"
#| fig-align: center

# Mid-parent height vs Child height plot
ggplot(GaltonFamilies, aes(x = midparentHeight, y = childHeight)) +
  geom_point(alpha = 0.5, color = "skyblue") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(
    title = "Heritability of Height: Mid-Parent vs Child",
    subtitle = "Data from Galton's study of families",
    x = "Mid-Parent Height (inches)",
    y = "Child Height (inches)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 12))
```

-   The positive slope of the trend line suggests a **direct correlation**, as mid-parent height increases, child height increases.
-   Some variability is observed around the line, indicating parental height is influential, other factors also affect child height.
-   The plot illustrates Galton’s insight, where parents with extremely tall or short height often have children whose heights are closer to the average.

# Conclusion

In summary, this plot illustrates that children’s heights generally relate to their parent's average height, reflecting genetic influence, while also highlighting the natural variation present in human growth.
